An Efficient Calculus for Generalised Parikh Images of Regular Languages

### 2022-07-06 main.tex
add marker
motivating example

Now assume that rather than the length (e.g. the count of the characters) we
want to find the whole Parikh image of the intersection of $\AcaOrBc{}$ and
$\SomethingCSomething{}$. In other words, now $f$ is a mapping to an object
(usually represented as a column vector) counting how often each character
occurs in a string. Since extracting values from a vector is tiresome, we will
use the shorthand notation $\TransitionCount{a}$ to refer to the number of
letters a that appear. E.g. $\TransitionCount{a}(\text{curious and curiouser
still}) = 1$. Note that we still have the property that
$\TransitionCount{o}(\text{homomorphism}) = \TransitionCount{o}(\text{homo}) +
\TransitionCount{o}(\text{morphism}) = 2 + 1 = 3$. If you treat the addition as
element-wise vector addition, this approach works for the vector version as
well.

For our constraint, we will use $\TransitionCount{a} > \TransitionCount{b}$,
that is find a count of characters in a string accepted by both automata where
there are more a:s than b:s. Since the underlying calculus is the same, we start immediately from Figure~\ref{fig:example-simplify-1}.

  We now need to relate the number of times a transition was taken to the
  character counts we are really interested in, as opposed to just the length in
  the previous example. Note that this, like the length calculus before, puts
  the two automata into contact with each other, before any product has been
  computed. Some consideration leads us to the definition that the number of
  occurrence of each character corresponds to the sum of usages of transitions
  containing it, e.g. relations of the kind $\TransitionCount{a} = 2 - 2l_c'$,
  and $\TransitionCount{b} = 2 l_b' + l_c'$.\footnote{For the attentive reader:
  in this example, we leave the catch-all transitions of $\SomethingCSomething$
  unconstrained since they will be taken care of by the product and would
  complicate the reasoning since we cannot yet know which character they will
  correspond to in the final product. In other words, this is a safe
  under-approximation that only limits our ability to reason before computing
  the product.}

This relation can be combined with the constraint that $\TransitionCount{a} >
\TransitionCount{b}$, that is that there are more instances in the accepted
string of letters a than b, to obtain $2 - 2l_c' > 2 l_b' + l_c'$. Combining this
equation with the requirement from $\SomethingCSomething$ that
$\TransitionCount{c} \geq 1$ and $\TransitionCount{c} = l_c + l_c'$ from
$\AcaOrBc$ gives $0 \geq l_b'$, and since we cannot use a transition a negative
number of times $l_b' = 0$, which gives the automaton in
Figure~\ref{fig:example-simplify-2}.

\begin{figure}[ht]
  \centering 
    \includegraphics[width=0.75\linewidth]{aca_or_bc_simplified_2}
    \caption{$\AcaOrBc{}$ after removing the now unused b-transition.}\label{fig:example-simplify-2}
  \end{figure}

  Plugging in the same inequality again on the same principle, we get $2(1 -
  l_c') > l_c' \iff 1 > l_c'$ since we are working with integers, and
  therefore that $l_c' = 0$. This gives us the even smaller automaton of
  Figure~\ref{fig:example-simplify-3}.

  \begin{figure}[ht]
    \centering 
      \includegraphics[width=0.75\linewidth]{aca_or_bc_simplified_3}
      \caption{$\AcaOrBc{}$ after further constraint propagation.}\label{fig:example-simplify-3}
    \end{figure}

This will produce a straightforward product since both automata now have the
same shape, where we essentially only need to expand the self-loops of
$\SomethingCSomething$ and reduce its range labels to only capture a:s, as seen
in Figure~\ref{fig:final-example}. Note that we got this result by only
performing linear inequality reasoning on the automata of the product versus the
number of times each transition would be used.

\begin{figure}[ht]
\centering
    \includegraphics[width=0.75\linewidth]{aca_or_bc_times_something_c_something}
    \caption{$\AcaOrBc{} \times \SomethingCSomething{}$ under the constraint
    $\TransitionCount{a} > \TransitionCount{b}$. Note the fresh variables
    introduced for the transitions (for most transitions the same one, since
    they are all linearly dependent).}\label{fig:final-example}
  \end{figure}

  \begin{figure}[ht]
\centering
    \includegraphics[width=0.75\linewidth]{original_product}
    \caption{$\AcaOrBc{} \times \SomethingCSomething{}$ as it would have
    appeared if we had computed it from the initial
    automata.}\label{fig:original-product}
  \end{figure}

To obtain the final Parikh image, we need to tie the inequalities for the
transitions of the automata of the product to the new transition counters we
introduced in the product automaton. The only interesting transition is the one
carrying the c, so we will choose that one for illustration. In that case, the
bridging equation is $r_2 = l_c \land r_2 = r_{\Sigma}$ (and for all others
$=1$), which gives the Parikh image $\exists_{r_2} \TransitionCount{a} = 1 \land
\TransitionCount{b} = 0 \land \TransitionCount{c} = r_2 + 1$, and after
existence-elimination using reasoning on lower bounds (e.g. $r_2 \geq 0$) we
arrive at $\TransitionCount{c} \geq 1$. For comparison, the $\AcaOrBc{} \times
\SomethingCSomething{}$ we would have obtained if we had eagerly computed the
product can be seen in Figure~\ref{fig:original-product}. This would have
introduced a significantly more complicated existentially quantified formula.

>>>>

Now assume that rather than the length (e.g. the count of the characters) we
want to find the whole Parikh image of the intersection of $\AcaOrBc{}$ and
$\SomethingCSomething{}$. In other words, now $f$ is a mapping to a 3-vector
counting how often each character occurs in a string. Since extracting values
from a vector is tiresome, we will use the shorthand notation
$\TransitionCount{a}$ to refer to the number of letters a that appear. E.g.
$\TransitionCount{a}(\text{curious and curiouser still}) = 1$. Note that we
still have the property that $\TransitionCount{o}(\text{homomorphism}) =
\TransitionCount{o}(\text{homo}) + \TransitionCount{o}(\text{morphism}) = 2 + 1
= 3$. If you treat the addition as element-wise vector addition, this approach
works for the vector version as well.

For our constraint, we will use $\TransitionCount{a} > \TransitionCount{b}$,
that is finding a count of characters in a string accepted by both automata where
there are more a:s than b:s. Since the underlying calculus is the same, we start immediately from Figure~\ref{fig:example-simplify-1}.

  We now need to relate the number of times a transition was taken to the
  character counts we are really interested in, as opposed to just the length in
  the previous example. For the $\Sigma$ labels, since we do not know which
  letter to use, we introduce a sum $r_{a1} + r_{b1} + r_{c1}$ per $\Sigma$
  transition.

  As in the previous example, we sum the occurrences for each transition to
  obtain the counts, here with the element-wise vector addition unpacked:
  \[
\MonoidElement =   \begin{bmatrix}
  r_{a1} + r_{a2} \\
  r_{b1} + r_{b2} \\
  r_{c1} + r_{c2} + 1 \\
  \end{bmatrix} =
  \begin{bmatrix}
    2 -2l_c' \\
    2l_b' + l_c' \\
    l_c + l_c' \\
    \end{bmatrix} =
    \begin{bmatrix}
      \TransitionCount{a} \\
      \TransitionCount{b} \\
      \TransitionCount{c}
      \end{bmatrix}
  \]

These equalities can be combined with the constraint that $\TransitionCount{a} >
\TransitionCount{b}$ to obtain $2 - l_c' > l_b'$. Since we have $1-l_c' \geq 0$
from one of the transition labels, we know $l_c' \geq 0$, and therefore that $1
> l_b' \implies l_b' = 0$. Plugging in the same inequality again on the same
principle, we get $2 - 2l_c' > l_c' \iff l_c' = 0$. This gives us the much
smaller automaton of \cref{fig:example-simplify-3}.

% \begin{figure}[t]
%   \centering 
%     \includegraphics[width=0.70\linewidth]{aca_or_bc_simplified_2}
%     \caption{$\AcaOrBc{}$ after removing the now unused b-transition.}\label{fig:example-simplify-2}
%     \Description[$\AcaOrBc$ without the returning transition from $B$ to $S$]{$\AcaOrBc$ without the returning transition from $B$ to $S$, which now makes the transition from $S$ to $B$ also have the same associated variable $l_c'$.}
%   \end{figure}

  \begin{figure}[t]
    \centering 
      \includegraphics[width=0.7\linewidth]{aca_or_bc_simplified_3}
      \caption{$\AcaOrBc{}$ after further constraint propagation.}\label{fig:example-simplify-3}
      \Description[$\AcaOrBc$ without the lower section.]{$\AcaOrBc$ with only its upper section left and the $a$ transitions now set to static 1.}
    \end{figure}

This will produce a straightforward product since both automata now have the
same shape, where we essentially only need to expand the self-loops of
$\SomethingCSomething$ and reduce its range labels to only capture a's, as seen
in \cref{fig:final-example}. Note that we got this result by only performing
linear inequality reasoning on the automata of the product versus the number of
times each transition would be used.

\begin{figure}[t]
\centering
    \includegraphics[width=0.75\linewidth]{aca_or_bc_times_something_c_something}
    \caption{$\AcaOrBc{} \times \SomethingCSomething{}$ under the constraint
    $\TransitionCount{a} > \TransitionCount{b}$. Note the fresh variables
    introduced for the transitions (for most transitions the same one, since
    they are all linearly dependent).}\label{fig:final-example} \Description[The
    final product, still a boring stick automaton with only one loop for an
    arbitrary number of c's.]{A product of the intermittent automata, which
    accepts \lstinline{ac.*ca}}
  \end{figure}

%   \begin{figure}[t]
% \centering
%     \includegraphics[width=0.75\linewidth]{original_product}
%     \caption{$\AcaOrBc{} \times \SomethingCSomething{}$ as it would have
%     appeared if we had computed it from the initial
%     automata.}\label{fig:original-product}
%   \end{figure}

To obtain the final Parikh image, we again need to tie the inequalities for the
transitions of the automata of the product to the new transition counters we
introduced in the product automaton. The only interesting transition is the one
carrying the c, so we will choose that one for illustration. In that case, the
bridging equation is $r_2 = l_c \land r_2 = r_{\Sigma}$ (and for all others
$=1$), which gives the Parikh image $\exists_{r_2} \TransitionCount{a} = 1 \land
\TransitionCount{b} = 0 \land \TransitionCount{c} = r_2 + 1$, and after
existence-elimination using reasoning on lower bounds (e.g. $r_2 \geq 0$) we
arrive at $\TransitionCount{c} \geq 1$.
